Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-12 thesis/thesis.tex
thesis: Update Deep Embedded Clustering section

\usepackage{apacite}
\usepackage{caption}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{picture}
\usepackage{subcaption}

@@

%---------------------------
% Título del trabajo y autor
%---------------------------
\title{Open Clusters Characterization in Gaia DR2 Using ML Algorithms}
\author{Carlos David Álvaro Yunta}
\date{13 de Septiembre de 2020}
\director{César Augusto Guzmán Álvarez}
\nombreciudad{Madrid, Spain}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/pm_melotte_22.pdf}
    \caption{Proper motions}
    \label{fig:pm_melotte_22}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/parallax_melotte_22.pdf}
    \caption{Parallax}
    \label{fig:pm_vec_melotte_22}
  \end{subfigure}
  \caption{Open Cluster Melotte 22 (M45)}
\end{figure}

@@

However, as shown in Figures \ref{fig:pm_ngc_2682} and \ref{fig:parallax_ngc_2682}, in general it is not as easy and
becomes necessary to consider other parameters such as distances, or even metallicity and age (derived from isochrone curves)
and requires photometric data from stars in the studied field.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/pm_ngc_2682.pdf}
    \caption{Proper motions}
    \label{fig:pm_ngc_2682}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/parallax_ngc_2682.pdf}
    \caption{Parallax}
    \label{fig:parallax_ngc_2682}
  \end{subfigure}
  \caption{Open Cluster NGC 2682}
\end{figure}

@@

% TODO: Hablar sobre los algoritmos de clustering que se van a emplear: KMeans, DEC

%% TODO: Concluir con una última sección de resumen de conclusiones, resumiendo las principales averiguaciones del estudio del dominio
% y cómo van a afectar al desarrollo específico del trabajo.

@@

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{../figures/openclust_catalogue.pdf}
  \caption{OpenClust Catalogue Distribution}
  \label{fig:OpenClustComplete}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{../figures/cluster_selection_tier1.pdf}
  \caption{OpenClust Catalogue Selection Distribution}
  \label{fig:OpenClustSelection}
\end{figure}

@@

The second container is just a \verb|postgres:12.4| image which loads an initial script when the database is not yet initialized for creating the
database structure. This container is used later as the main database for data analysis.

@@

These containers are orchestrated with a \verb|docker-compose.yml| file which automatilly launches one instance of each container.

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_pm_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_parallax_melotte_22.pdf}
      \caption{N clusters = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_parallax_melotte_22.pdf}
      \caption{N clusters = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_parallax_melotte_22.pdf}
      \caption{N clusters = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

@@

Our initial approach allows us to find some groups that potentially contain the OC. However we need now a method to merge clusters
by migrating stars from one group to another. So we can end up with the minimum number of clusters while preserving the open cluster.

@@

DEC is divided in two stages. In the first one and \emph{deep autoencoder} is pretrained to extract information from data and generate the
the embedding vectors. In the second stage, a parameter optimization is performed, iterating between computing an auxiliary target distribution
and minimizing the Kullback-Leiber (KL) divergence \cite{kullback1951information} to it.

This model uses K-Means as the starting point, since it is used to create an initial clustering and then the DEC model is used to refined it.

@@

%% TODO: Las tecnologías empleadas para el desarrollo de las herramientas,
% la organización del piloto, cómo ha transcurrido el experimento

%% TODO: Añadir nuevo capítulo - Descripción de los resultados

>>>>

\usepackage{apacite}
\usepackage{caption}
\usepackage{eso-pic}
\usepackage{float}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{picture}
\usepackage{subcaption}

@@

%---------------------------
% Título del trabajo y autor
%---------------------------
\title{Open Clusters Characterization in Gaia DR2 Using ML Algorithms}
\author{Carlos David Álvaro Yunta}
\date{12th of November, 2020}
\director{César Augusto Guzmán Álvarez}
\nombreciudad{Madrid, Spain}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/pm_melotte_22.pdf}
      \caption{Proper motions}
      \label{fig:pm_melotte_22}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/parallax_melotte_22.pdf}
      \caption{Parallax}
      \label{fig:pm_vec_melotte_22}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster Melotte 22 (M45)}
\end{figure}

@@

However, as shown in Figures \ref{fig:pm_ngc_2682} and \ref{fig:parallax_ngc_2682}, in general it is not as easy and
becomes necessary to consider other parameters such as distances, or even metallicity and age (derived from isochrone curves)
and requires photometric data from stars in the studied field
\footnote{All figures have been sampled in order to reduce the document size, but are still a faithful representation of the original data}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/pm_ngc_2682.pdf}
      \caption{Proper motions}
      \label{fig:pm_ngc_2682}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/parallax_ngc_2682.pdf}
      \caption{Parallax}
      \label{fig:parallax_ngc_2682}
    \end{subfigure}
  \end{subfigure}
  \caption{Open Cluster NGC 2682}
\end{figure}

@@

% TODO: Hablar sobre los algoritmos de clustering que se van a emplear: K-Means, DEC

% TODO: Concluir con una última sección de resumen de conclusiones, resumiendo las principales averiguaciones del estudio del dominio
% y cómo van a afectar al desarrollo específico del trabajo.

@@

% TODO: Hay que desarrollar más los objetivos

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/openclust_catalogue.pdf}
  \caption{OpenClust Catalogue Distribution}
  \label{fig:OpenClustComplete}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/cluster_selection_tier1.pdf}
  \caption{OpenClust Catalogue Selection Distribution}
  \label{fig:OpenClustSelection}
\end{figure}

@@

Downloader image build instructions are available at \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/downloader.Dockerfile}{downloader.Dockerfile}
and latest image is available for pulling from \href{https://github.blog/2020-09-01-introducing-github-container-registry/}{GitHub Container Registry}:

\begin{minted}{sh}
  docker pull ghcr.io/cdalvaro/gaia-downloader:latest
\end{minted}

This image is automatically built with \href{https://github.com/features/actions}{GitHub Actions} on every push to the \verb|main| branch.

The second container is just a \verb|postgres:12.4| image which loads an initial script when the database is not yet initialized for creating the
database schema. This container is used later as the main database for data analysis.

@@

These containers are orchestrated with a \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/docker-compose.yml}{docker-compose.yml}
file which automatilly launches one instance of each container.

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_pm_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_pm_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n2_parallax_melotte_22.pdf}
      \caption{N clusters = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n5_parallax_melotte_22.pdf}
      \caption{N clusters = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_n8_parallax_melotte_22.pdf}
      \caption{N clusters = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

@@

Our initial approach allows us to find some groups that potentially contain the desired OC. However we need now a method for merging clusters
by migrating stars from one group to another. So we can end up with the minimum number of clusters while preserving the open cluster.

Since we don't have a labeled dataset which allows as to train a supervised model, we are required to use an unsupervised self-trained model.

@@

As we can see in Figure \ref{fig:dec_melotte_22}, the DEC model is able to merge similar clusters leaving alone the OC. Which is exactly what we are looking for.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_pm_melotte_22.pdf}
      \caption{Proper motions}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_isochrone_melotte_22.pdf}
      \caption{Isochrone curve}
    \end{subfigure}
  \end{subfigure}
  \caption{DEC model applied to Melotte 22}
  \label{fig:dec_melotte_22}
\end{figure}

The implementation of this model is available in \verb|cdalvaro.ml.DEC| and it is developed with the Keras framework.

The model is composed by a \emph{deep autoencoder}, a \emph{deep encoder} and a \emph{clustering layer}.

Encoders are used to transform the input data into a latent space using a non-linear mapping function $f_{\theta} : X \rightarrow Z$.

Although in this work we are managing a small number of features, this latent space allows us to reduce the number of features, avoiding
this way the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions. Then, the encoder is used to transform input data to the
latent space $Z$. Once the data has been transformed, a K-Means clusterer is used to make an initial clustering of the data.

With that initial configuration, the model iterates alternating between computing an auxiliary target distribution (Soft Assignment)
and minimizing the Kullback-Leiber (KL) divergence \cite{kullback1951information} to it. This unsupervised algorithm allows us to improve
the clustering.

In the soft assignment stage, the \emph{Student's t-distribution} is used as a kernel to measure the similarity between the embedded points
and the cluster centroid. While in the KL divergence minimization the algorithm iteratively refines clusters by learning from their high
confidence assignments with the help of an auxiliary target distribution. The model is trained by matching the soft assignment to the target
distribution. The choice of this target distribution is crucial for DEC's performance. In this work we have taken the target distribution
from DEC's original paper \cite{xie2016unsupervised}, which is defined as follows:

\begin{equation}
  p_{ij} = \frac{q^{2}_{ij} / f_{j}}{\sum_{j'}q^{2}_{ij'}/f_{j'}}
\end{equation}

Inside \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks}{src/notebooks},
there is a notebook called
\href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks/cluster_characterization.ipynb}{cluster\_characterization.ipynb}
that illustrates how DEC model process a region of stars trying to characterize the OC hidden inside that region.

@@

% TODO: Las tecnologías empleadas para el desarrollo de las herramientas,
% la organización del piloto, cómo ha transcurrido el experimento

% TODO: Añadir nuevo capítulo - Descripción de los resultados
